---
title: "Article S3: Clustering and isolation by distance"
output:
  pdf_document:
    toc: yes
    df_print: kable
  html_document:
    code_folding: hide
    df_print: paged
    number_sections: yes
    theme: readable
    toc: yes
    toc_float:
      collapsed: no
---

In this notebook we will do a few simple population genetics analyses to ttest the use of ddRAD libraries prepared with Multiple Displacement Amplification (MDA)

Let's start by loading packages

```{r}
library(adegenet)
library(BEDASSLE)
library(fossil)
library(dplyr)
library(RColorBrewer)
library(ggplot2)
library(broom)
library(data.table)
library(ggthemes)
library(gridExtra)
library(vegan)
library(knitr)

knitr::opts_chunk$set(warning=FALSE)
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
```

Now let's load the data:

```{r}

taxa = list.files('./',pattern = '.*gen$') %>%
  gsub(pattern = '_filtered.*$',replacement = '')

genpops = sapply(taxa,function(x){
  paste(x,'.*gen$',sep='') %>%
  list.files(path = './') %>%
  read.genepop(ncode=3,quiet=FALSE)
})

sample_info = read.csv('sample_info_new_WGA.csv')

pop_locations = read.csv('pop_locations.csv')

```

Now let's use DAPC to find clusters. We manually selected number of clusters according to variation in BIC for different number of clusters. 
Basically, we selected the number of groups right before a sharp increase in BIC. In the case of **Anchylorhynchus**, number of clusters was ever decreasing. We selected 4 clusters because that is the point in which the curve seems to start flattening.

This resulted in the following number of clusters:

```{r}
nclusters = c('Anchylorhynchus' = 4,
              'Andranthobius' = 1,
              'C_impar' = 3,
              'M_bondari' = 2,
              'M_ypsilon' = 2)
```


The following runs find.clusters (we first ran interactively to get the number of clusters above). To run interactively, one needs to change `choose.n.clust` to `FALSE`.
```{r}
dapcs = lapply(taxa,function(x){
  tryCatch(expr = {
    re = find.clusters(tab(genpops[[x]],freq=T,NA.method='mean'),choose.n.clust = FALSE,pca.select='percVar',perc.pca=75,n.clust = nclusters[x])
    return(re)
    },
    error=function(e){
      samps = rownames(genpops[[x]]$tab)
      outvec = rep(1,length(samps))
      names(outvec) = samps
      re = list(grp=outvec)
      return(re)
    }
  )
  }
  )
names(dapcs) = taxa
```


Now let's plot make plots PCAs including genetic clusters and local populations. We will save plots in a list and make a multipnale plot later.

```{r}
titles = c('Anchylorhynchus' = 'Anchylorhynchus',
           'Andranthobius' = 'Andranthobius',
           'C_impar' = 'Celetes impar',
           'M_bondari' = 'Microstrates bondari',
           'M_ypsilon' = 'Microstrates ypsilon')
plots=list()

for (taxon in names(genpops)){
  gendata = genpops[[taxon]]
  
  imputed = tab(gendata, freq=TRUE, NA.method = 'mean')
  pca1 = dudi.pca(df = imputed, scale = FALSE, scannf = F, nf=20)
  
  xrange = range(pca1$li$Axis1) 
  yrange = range(pca1$li$Axis2)
  
  xrangeplot = xrange + c(-1,1)*(xrange[2]-xrange[1])*0.3
  yrangeplot = yrange + c(-1,1)*(yrange[2]-yrange[1])*0.3
  
  pc1_var = 100*pca1$eig[1]/sum(pca1$eig)
  pc2_var = 100*pca1$eig[2]/sum(pca1$eig)
  
  localities = gsub('^.*_','',rownames(pca1$li)) %>%
    factor
  
  samples = gsub('_.*$','',rownames(pca1$li))
  
  MDA = sample_info %>%
    filter(samplename_ipyrad %in% samples) %>%
    select(WGA) %>% 
    unlist %>%
    factor(ordered=T,levels=c('TRUE','FALSE'))
  
  plot_df = data.frame(sample = samples,
                       locality = localities,
                       cluster = dapcs[[taxon]]$grp,
                       MDA = MDA,
                       PC1 = pca1$li[,1],
                       PC2 = pca1$li[,2])
  
  #Some groups have few data points, and stat_ellipse needs at least 3 points.To make ellipses consistently, we will double the number of points and put a small jitter. The idea is to simply enclose populations, not to be statistically accurate.
  
  ellipse_df = rbind(plot_df,plot_df) %>%
    mutate(PC1 = PC1+rnorm(n = length(PC1),mean = ,sd=0.1),
           PC2 = PC2+rnorm(n = length(PC2),mean = ,sd=0.1)) %>%
    select(locality,PC1,PC2)
  
  centroids = plot_df %>%
    group_by(cluster) %>%
    summarise(cent.PC1 = mean(PC1),
              cent.PC2 = mean(PC2)) %>%
    right_join(plot_df)
  
  if (taxon == 'Anchylorhynchus'){
    ellipses = NULL
  }
  else {
    ellipses = stat_ellipse(aes(x=PC1, y=PC2, group=locality),linetype='dashed',data=ellipse_df,type='t',color='grey30') 
  }
    p = ggplot(plot_df) +
    ellipses +
    geom_segment(aes(x=PC1, y=PC2, xend=cent.PC1,yend=cent.PC2), data = centroids) +
    geom_point(aes(x=PC1, y=PC2, color=MDA)) +
    xlab(paste('PC1 (',sprintf('%.1f',pc1_var),'%)',sep='')) +
    ylab(paste('PC2 (',sprintf('%.1f',pc2_var),'%)',sep='')) +
    theme_tufte() +
    theme(panel.border = element_rect(colour = "black", fill=NA)) +
    scale_colour_manual(values = brewer.pal(n=3, name = 'RdYlBu')[c(1,3)], name = '',labels = c('FALSE'='gDNA','TRUE'='MDA')) +
    ggtitle(bquote(italic(.(titles[taxon])))) +
    scale_y_continuous(labels=function(x)sprintf('%1.1f',x))
  
  plots[[taxon]] = p
  
}
```

Now let's plot all of them together:

```{r}
#extract legend. From:
#https://github.com/hadley/ggplot2/wiki/Share-a-legend-between-two-ggplot2-graphs
g_legend<-function(a.gplot){
  tmp <- ggplot_gtable(ggplot_build(a.gplot))
  leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
  legend <- tmp$grobs[[leg]]
  return(legend)}

my_legend = g_legend(plots[[1]])
  
p = grid.arrange(plots[[1]] + theme(legend.position='none'),
                         plots[[2]] + theme(legend.position='none'),
                         plots[[3]] + theme(legend.position='none'),
                         plots[[4]] + theme(legend.position='none'),
                         plots[[5]] + theme(legend.position='none'),
                         my_legend, ncol=2)

print(p)

ggsave(filename = 'fig_DAPC.pdf',plot = p,device = 'pdf',path = 'plots/',width = 6.5, height=8, useDingbats=F)
```

Now we will look at the effect of MDA on isolation by distance. We will use only  **C. impar** only **Anchylorhynchus** is complicated and includes many species, while the other taxa have too few populations

We will start by defining a function that calculates pairwise FSTs. The standard function in hierfstat seems to calculate FST wrongly. I am using the function in BEDASSLE, which implements the "ratio of averages" as defined by Bhatia G., Patterson N., Sankararaman S., Price AL. 2013. Estimating and interpreting FST: The impact of rare variants. Genome Research 23:1514–1521. DOI: 10.1101/gr.154831.113.

```{r}
get_pairwise_Fst = function(genmat,grp){
  grp = factor(grp)
  if (length(levels(grp)) > 1){
    
    allele.counts = apply(genmat@tab,2,function(x){
      pops = grp
      counts = tapply(x,pops,function(y){
        sum(y,na.rm = T)
      })
      return(counts)
    })
    sample.sizes = apply(genmat@tab,2,function(x){
      pops = grp
      sizes = tapply(x,pops,function(y){
        2*sum(!is.na(y))
      })
      return(sizes)
    })
    
    Fst = calculate.all.pairwise.Fst(allele.counts,sample.sizes)
    colnames(Fst) = rownames(Fst) = levels(grp)
    #print(Fst)
    return(tidy(as.dist(Fst,upper = FALSE)))
  } else {
    print('Only one pop')
    return(matrix())
  }
}
```

Now let's calculate the pairwise geographical distances between populations
```{r}

geodist = earth.dist(pop_locations[c('lon','lat')], dist = FALSE)
rownames(geodist) = colnames(geodist) = pop_locations$population

geodist = geodist[sort(rownames(geodist)),sort(rownames(geodist))] %>%
  as.dist(upper=TRUE) %>%
  tidy() %>%
  transmute(population1=item1,
            population2=item2,
            geo_distance=distance)

head(geodist,n = 20)

```



Now let's calculate FSTs for **C. impar** (all samples, only MDA and only gDNA) and save results in a table
```{r}
FST = plyr::ldply(c('all','MDA','gDNA'), function(y){
    samples =  rownames(genpops$C_impar@tab)
    pops = gsub('^.+_','',samples)
    samples = gsub('_.+$','',samples)
    
    if (y=='all'){
      return(get_pairwise_Fst(genpops$C_impar,pops) %>% mutate(method = y))
    }
    else if (y == 'MDA'){
      samples = sample_info %>%
        filter(WGA == TRUE, samplename_ipyrad %in% samples) %>%
        tidyr::unite('sample_pop',samplename_ipyrad, population) %>%
        select(sample_pop) %>%
        unlist
      pops = gsub('^.+_','',samples)
      return(get_pairwise_Fst(genpops$C_impar[samples],pops)%>% mutate(method = y))
    }
    else {
      samples = sample_info %>%
        filter(WGA == FALSE, samplename_ipyrad %in% samples) %>%
        tidyr::unite('sample_pop',samplename_ipyrad, population) %>%
        select(sample_pop) %>%
        unlist
      pops = gsub('^.+_','',samples)
      return(get_pairwise_Fst(genpops$C_impar[samples],pops)%>% mutate(method = y))
    }
    
    
    
  })

FST = FST %>% transmute(method, population1=item1, population2=item2, FST=distance)

FST

```

Now let's join geographical distance and linearize FST:

```{r}
FST_plot = FST %>% left_join(geodist) %>% mutate(FST = (FST)/(1-FST))
FST_plot
```

Now let's plot:
```{r}
p = ggplot(FST_plot,aes(x=geo_distance, y = FST, color=method)) +
  geom_point() +
  geom_smooth(method=lm,se = FALSE) +
  theme_tufte() +
  theme(panel.border = element_rect(colour = "black", fill=NA)) +
  scale_colour_manual(values = c('black',brewer.pal(n=3, name = 'RdYlBu')[c(3,1)]), name = '') +
  ylab(bquote('F'['ST']/(1-'F'['ST']))) +
  xlab('Geographical distance (km)')

print(p)

ggsave(filename = 'fig_Mantel.pdf',plot = p,device = 'pdf',path = 'plots/')
  
```

Finally, let's do Mantel tests and see if results are different between methods:

```{r}
for (method in c('all','gDNA','MDA')){
  
  FST_mat = FST_plot %>%
    filter(!!method == method) %>%
    dcast(formula=population2~population1,value.var='FST') %>%
    (function(x){rownames(x) = x[[1]]; return(x[-1])}) %>%
    as.matrix 
  
  pops = base::union(rownames(FST_mat),colnames(FST_mat)) %>%
    sort
  
  FST_dist = matrix(NA,nrow=length(pops),ncol=length(pops),dimnames = list(pops,pops))
  
  FST_dist[t(combn(pops,2))] = FST_mat[t(combn(pops,2))]
  
  FST_dist = FST_dist %>%
    t %>%
    as.dist
  
  
  
  geo_mat = FST_plot %>%
    filter(!!method == method) %>%
    dcast(formula=population2~population1,value.var='geo_distance') %>%
    (function(x){rownames(x) = x[[1]]; return(x[-1])}) %>%
    as.matrix
   
  geo_dist = matrix(NA,nrow=length(pops),ncol=length(pops),dimnames = list(pops,pops))
  
  geo_dist[t(combn(pops,2))] = geo_mat[t(combn(pops,2))]
  
  geo_dist = geo_dist %>%
    t %>%
    as.dist
  
  mantel_res = mantel(geo_dist,FST_dist)
  print(method)
  print(mantel_res)
}
```

This lists number of samples per population Table S2:
```{r}
for (taxon in taxa){
  print(taxon)
  print(sample_info[sapply(sample_info$samplename_ipyrad,function(x)any(grepl(x,rownames(genpops[[taxon]]@tab)))),] %>% select(population,WGA) %>% table)
}

```

